A Mechanised Library for Generic (Co)Programming

### 2024-03-15 paper/paper.tex
Updates

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of
\emph{setoids}, types with an associated equivalence relation, and considering
only functions that respect the associated equivalences, or \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation. We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code. We use \coq{=e} to denote the equivalence
relation of a setoid. By default, we associate every Coq type with the standard
propositional equality, unless a different equivalence is specified (we allow
overlapping instances, and Coq's propositional equality takes the lowest
priority).
\begin{coqcode}
Reserved Notation "f =e g" (at level 70, no associativity).
Class setoid A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Structure morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
We rely on Coq's type class mechanism to fill in the necessary equivalence
relations. Coq's code extraction mechanism will erase any occurrence
of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be extracted to
the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
from \coq{A ~> B} to \coq{A -> B}.
On top of this, we define basic function composition and identity functions:
\begin{coqcode}
Notation "f \o g" = (comp f g).
Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
Definition id : A ~> A := ...
\end{coqcode}
Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape * Pos ~> bool }.
Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
however, we chose to use Coq's propositional equality, since these would leads
to simpler code. We explain why in Section~\ref{sec:algebras}, and the
mechanisation of initial algebras of container extensions.
%The definitions in
%Figure~\ref{fig:poly} are redefined to account for these changes in a
%straightforward way, i.e. we prove that validity predicates are indeed proper
%morphisms.

This definition leads to the well-known ``setoid hell'', which we mitigate by
providing tactics and notations to automatically discharge proofs of
\coq{app_eq} for morphisms, whenever the types use the standard propositional
equality, or a combination of propositional and extensional equality. However,
our compositional approach allows us to build morphisms by plugging in other
morphisms to our combinator. In our framework, our expectation is that the
user-provided functions remain small, with relatively straightforward proofs of
\coq{app_eq}.

@@

We prove the isomorphisms of polynomial types and the equivalent (redefined)
container extensions. As an example, we discuss (informally) the isomorphisms
of pairs with their equivalent container extensions. Suppose that we
know that \coq{App F X} is isomorphic to \coq{A}, and
\coq{App G X} is isomorphic to \coq{B}. Then we can
show that \coq{App (Prod F G) X} is isomorphic to
\coq{A * B}.  If we have an element of type 
\coq{App (Prod F G) X},
using the \coq{inl} position, we can obtain
\coq{App F X}.  Similarly, using \coq{inr}, we can
obtain \coq{App G X}. Since these are the only two valid positions
in the shape of pairs, we have finished. It is now sufficient to use the
isomorphisms of \coq{App F X} and \coq{App G X} to
obtain \coq{A * B}. Similarly if we have \coq{A * B},
we can first use the isomorphisms of 
\coq{A} and
\coq{B} to obtain
\coq{App F X * App G X}, and then construct the necessary container
extension. Given \coq{p_inl}/\coq{p_inr} that act as
\coq{inl/inr} on valid product positions, the relevant
functions that define the isomorphism are:
\begin{coqcode}
Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
  ({| shape := shape (fst x); cont := fun e => cont x (e_inl e) |}, 
   {| shape := shape (snd x); cont := fun e => cont x (e_inr e) |}).
Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
{| shape := (shape (fst x), shape (snd x));
   cont := fun e => match val e with 
                    | inl p => fun H => cont (fst x) (ValidPos p H)
                    | inr p => fun H => cont (snd x) (ValidPos p H)
                    end (IsValid e) |}.
\end{coqcode}
Proving that these functions indeed form an isomorphism is straightforward,
using the fact that \coq{Prod} containers only have two valid positions.

\section{Formalising Recursion Schemes}
Recursion schemes provide an abstract way to consume and generate data. We know
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise recursive coalgebras (Section~\ref{sec:rec-coalgebras}) and then we
put together these notions to formalise hylomorphisms (Section~\ref{sec:hylo}).

\subsection{Algebras and Catamorphisms}
\label{sec:algebras} An algebra is a set \coq{X} together with some operations
on it\footnote[1]{Sometimes and algebra comes equipped with an equational
theory, e.g. the theory of monoids, but here we focus on algerbraically free
algebras}.  For example, give such a type \coq{X} we can define the monoid
operations as \coq{u : unit -> X} for unit and \coq{m : X * X -> X} for
multiplication. Notice that \coq{FX = unit + X * X} is a functor on \coq{X}
which means we can equivalently describe these two maps as a single one of
type \coq{F X -> X} which we call an algebra for the functor \coq{F}
satisfying the unit and associative laws of the monoid.  In this work we will be
using algebras to describe the operations of a data type and so we will not be
needing additional equations on these operations.

Since in this work we will be considering only polynomial functors we need to
formalise algebras in terms of containers and containers extensions. Given a
type `\coq{A}` and a container \coq{F} an algebra for \coq{F} over the carrier
`\coq{A}` is a map of the following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}

Now, the \emph{easiest} and most effective way to turn any type $A$ into a
monoid is to define it as the least solution to the equation
\begin{coqcode}
Free A = A + unit + Free A * Free A
\end{coqcode}
with \coq{Free : Type -> Type}. In Coq, this is the least fixed-point of the
functor \coq{unit + A + X * X} on \coq{X}.  This is isomorphic to the type of
lists over \coq{A} hence it is the least fixed-point for the
functor \coq{unit + A * X} as well. This is the least ``set'' generated by the
constructors \coq{Empty : unit -> List A} and \coq{Cons : A * List A -> List A}.
These constructors give an algebra \coq{LFix_in : unit + A * List A -> List A}
which has and inverse which we call \coq{LFix_out}.

We now formalise this idea using container extensions.  For a container
extension a data type is formalised as an inductive data type parametrised by
the functor represented by \coq{F}:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
% To generalise recursion schemes we abstract the shape of the data using functors.
% Assuming a base functor \coq{F} a catamorphism is then the unique
% solution to the equation
% \begin{minted}{coq}
% cata alg = alg . fmap (cata alg) . InOp
% \end{minted}
% which does precisely what we described in the previous paragraph.

% where $F$ is a \emph{base functor}, $Ff$ is the \emph{functorial action} which
% can be thought of the program $\operator{fmap}\; f$, $a$ is a program of type
% $F A \to A$ (an $F$-algebra),  $c$ is a program of type $C \to FC$ (a
% $F$-coalgebra), $\InOp$ is the isomoprhism on the least fixed-point of $F$,
% namely $\InOp : \mu F \to F\mu F$, and $\OutOp$ is the isomorphism on the
% greatest fixed-point of $F$, namely $F \nu F \to \nu F$. When the unique
% solutions to the above equations exist we denote them as $\cata{a}$ and
% $\ana{c}$ respectively.

%%\dcas{Boring stuff}
%The definition of \coq{LFix} in terms of \coq{App} requires us to define our
%own custom induction scheme for initial algebras:
%\begin{coqcode}
%Lemma LFix_ind [P : LFix -> Prop]
%  : (forall x : App F LFix,
%        (forall e : Pos (shape x), P (cont x e)) ->
%        P (LFix_in x))
%    -> forall l : LFix, P l.
%\end{coqcode}
We define smart constructors for the isomorphism of least fixed points
\coq{LFix_out}.
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F
l_out : LFix F ~> App F (LFix F)
\end{coqcode}

The most simple and well-known recursion scheme is the \emph{fold} or
\emph{catamorphism} which consumes data. First a fold \emph{destructs} a data
type using \coq{LFix_out}, then calls itself recursively and then folds the
results of the recursive call using an \coq{F}-algebra.
\begin{coqcode}
Definition cata_f (alg : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      alg (MkApp sx (fun e => f (kx e)))
  end.
\end{coqcode}
It is easy to show that this function is a morphism of \coq{F}-algebras. In fact,
it is possible to define it as a map of the following type:
\begin{coqcode}
Alg F A ~> LFix ~> A
\end{coqcode}

We prove that catamorphisms satisfy the universal law, that is that there is a
unique morphism of \coq{F}-algebras from the least fixed-point of \coq{F}:
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}

\subsection{Coalgebras and Anamorphisms}
While folds \emph{destruct} data, their dual, the unfolds, \emph{construct}
data. In particular, given a seed value and an \emph{observation}, the unfold
corecursively  ``runs'' the observations starting from the initial seed.

% Consider the transition system below composed by three states \coq{x1}, \coq{x2}
% and \coq{x3} with the transitions labelled by the set of labels \coq{\{a, b\}} and define the map \coq{c} such that it implements the transition system below
% \[\begin{tikzcd}
%     {x_1} & {x_2} & {x_3} \arrow["b"', loop, distance=2em, in=35, out=325]
% 	\arrow["a", from=1-1, to=1-2]
% 	\arrow["b", from=1-2, to=1-3]
% \end{tikzcd}\]

The seed function can be modelled by a coalgebra. For example, for a type of
states \coq{X} and an alphabet type \coq{L} we can define a (labelled)
transition system (LTS) on \coq{X} as a function \coq{c : X -> L * X}
implementing the \emph{transition} map. In particular, for a
state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.
Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
carrier \coq{X} is a map \coq{X -> FX}. In the previous example \coq{c} is
an \coq{F}-coalgebra on the carrier \coq{X} with \coq{FX = L * X}.
In our development we use the following notation for coalgebras:
\begin{coqcode}
Notation Coalg F A := (A ~> App F A) (only parsing).
\end{coqcode}

Now an anamorphism takes such a seed function and runs it while collecting all
the observable behaviours. In order to do so we define the type of infinite
streams of labels which is the greatest fixed-point for the functor \coq{A * X}.
We define this in Coq as follows:
\begin{coqcode}
CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
\end{coqcode}
where \coq{GFix_in}  and \coq{GFix_out} are the two inverses witnessing the
isomorphism. Additionally, we define smart constructors for the isomorphism of greatest fixed points
\begin{coqcode}
g_in : App F GFix ~> GFix
g_out : GFix ~> App F GFix
\end{coqcode}

What witnesses the fact that this is the greatest fixed-point is the universal
property of the anamorphism. For a given \coq{F}-coalgebra this is the unique
map that runs the seed function and composes the information from the first step
the result

\begin{coqcode}
Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
: f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
Equivalently, we could have define an anamorphism as the
unique \coq{F}-coalgebra homomorphism, that is a map of the following type
\begin{coqcode}
Coalg F A ~> A ~> GFix
\end{coqcode}

\subsection{Recursive Coalgebras and Hylomorphism}
\label{sec:rec-coalgebras}
Hylomorphism capture the concept of \emph{divide-and-conquer} algorithms.  First
the input is destructured (\emph{divide}) in smaller parts. This operation may
or may not be structural. Next the results from the subparts are computed
recursively and finally the results are composed together (\emph{conquer}) by
means of an algebra.

@@

Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
unique solution to the equation
\begin{equation}
  \label{eq:hylo}
  f = a \circ F\; f \circ c
\end{equation}

However this solution does not necessarily exists for an arbitrary pair of
algebras and coalgebras. In fact, this definition cannot be accepted by Coq,
since there is no structurally smaller argument to do recursion on and there is
no corecursive definition that is guaranteed to be syntactically guarded -- as
required by Coq -- due to the use of an arbitrary algebra $a$. Furthermore,
solutions to the hylomorphism equation are \emph{not} guaranteed to be unique,
or to even exist. To address these issues, we mechanise \emph{recursive
hylomorphisms} which are guaranteed to have a unique solution to the
hylomorphism equation. These are hylomoprhisms where the coalgebra is
\emph{recursive}.

Recursive coalgebras are coalgebras whose anamorphism produces finite data which
what the following predicate captures:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
Recursive anamorphisms are implemented in Coq recursively on the structure of
the proof for the type (\coq{RecF}) as follows:
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A)
  : forall (x : A), RecF c x -> B := fix f x H
  := match c x as h0 return
       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with
     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
We again use \coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls. As we did with catamorphisms and recursive anamorphisms, we
prove that \coq{hylo_def} is respectful, and use this proof to built the
corresponding higher-order proper morphism:
\begin{coqcode}
Alg F B ~> RCoalg F A ~> A ~> B
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to the
hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
Hylomorphisms fusion falls out from the universal property:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
\end{coqcode}
% Qed.
%   rewrite hylo_univ, fmap_comp, compA, <- E2,
%           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.

>>>>

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of
\emph{setoids}, types with an associated equivalence relation, and considering
only functions that respect the associated equivalences, or \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation. We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code. We use \coq{=e} to denote the equivalence
relation of a setoid. By default, we associate every Coq type with the standard
propositional equality, unless a different equivalence is specified (we allow
overlapping instances, and Coq's propositional equality takes the lowest
priority).
%\begin{coqcode}
%Reserved Notation "f =e g" (at level 70, no associativity).
%Class setoid A : Type :=
%  MkEquiv
%    { eqRel : A -> A -> Prop;
%      e_refl : forall x, eqRel x x;
%      e_sym : forall x y, eqRel x y -> eqRel y x;
%      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
%    }.
%Notation "f =e g" := (eqRel f g).
%\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represent \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Structure morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
We rely on Coq's type class mechanism to fill in the necessary equivalence
relations. Coq's code extraction mechanism will erase any occurrence
of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be extracted to
the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
from \coq{A ~> B} to \coq{A -> B}.
On top of this, we define basic function composition and identity functions:
\begin{coqcode}
Notation "f \o g" = (comp f g).
Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
Definition id : A ~> A := ...
\end{coqcode}
Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Class Cont `{setoid Shape} (Pos : Type) := { valid : Shape * Pos ~> bool }.
Inductive AppR `{F : Cont Sf Pf} `{setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
however, we chose to use Coq's propositional equality, since these would leads
to simpler code. We explain why in Section~\ref{sec:algebras}, and the
mechanisation of initial algebras of container extensions.
%The definitions in
%Figure~\ref{fig:poly} are redefined to account for these changes in a
%straightforward way, i.e. we prove that validity predicates are indeed proper
%morphisms.

This definition leads to the well-known ``setoid hell'', which we mitigate by
providing tactics and notations to automatically discharge proofs of
\coq{app_eq} for morphisms, whenever the types use the standard propositional
equality, or a combination of propositional and extensional equality. However,
our compositional approach allows us to build morphisms by plugging in other
morphisms to our combinators. In our framework, our expectation is that the
user-provided functions remain small, with relatively straightforward proofs of
\coq{app_eq}.

@@

We prove the isomorphisms of polynomial types and the equivalent 
container extensions. As an example, we discuss (informally) the isomorphisms
of pairs with their equivalent container extensions. Suppose that we
know that \coq{App F X} is isomorphic to \coq{A}, and
\coq{App G X} is isomorphic to \coq{B}. Then we can
show that \coq{App (Prod F G) X} is isomorphic to
\coq{A * B}.  If we have an element of type 
\coq{App (Prod F G) X},
using the \coq{inl} position, we can obtain
\coq{App F X}.  Similarly, using \coq{inr}, we can
obtain \coq{App G X}. Since these are the only two valid positions
in the shape of pairs, we have finished. It is now sufficient to use the
isomorphisms of \coq{App F X} and \coq{App G X} to
obtain \coq{A * B}. Similarly if we have \coq{A * B},
we can first use the isomorphisms of 
\coq{A} and
\coq{B} to obtain
\coq{App F X * App G X}, and then construct the necessary container
extension. Given \coq{p_inl : Pos l -> Pos (l * r)} (resp. \coq{p_inr}) that
act as \coq{inl} (resp.\ \coq{inr}) on product positions, and
\coq{case_pos : (Pos l -> X) -> (Pos r -> X) -> Pos (l * r) -> X} that pattern
matches on the product positions, the functions that witness the isomorphism
are:
\begin{coqcode}
Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
  ({| shape := shape (fst x); cont := fun e => cont x (p_inl e) |}, 
   {| shape := shape (snd x); cont := fun e => cont x (p_inr e) |}).
Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
{| shape := (shape (fst x), shape (snd x));
   cont := case_pos (cont (fst x)) (cont (snd x)) |}.
\end{coqcode}
Proving that the composition of these functions is the identity is
straightforward using the fact that \coq{Prod} containers only have two valid
positions.

\section{Formalising Recursion Schemes}
Recursion schemes provide an abstract way to consume and generate data. We know
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
these notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:rec-coalgebras}).

\subsection{Algebras and Catamorphisms}
\label{sec:algebras} An algebra is a set \coq{X} (the \emph{carrier} of the
algebra) together with some operations on it\footnote[1]{Sometimes and algebra
comes equipped with an equational theory, e.g. the theory of monoids, but here
we focus on algebraically free algebras}.  For example, given such a type
\coq{X} we can define the monoid operations as \coq{u : unit -> X} for unit and
\coq{m : X * X -> X} for multiplication. Notice that \coq{F X = unit + X * X}
is a functor on \coq{X} which means we can equivalently describe these two maps
as a single one of type \coq{F X -> X} which we call an algebra for the functor
\coq{F} satisfying the unit and associative laws of the monoid.  In this work
we will be using algebras to describe the operations of a data type and so we
will not be needing additional equations on these operations.

Since in this work we will be considering only polynomial functors we need to
formalise algebras in terms of containers and containers extensions. Given a
type \coq{A} and a container \coq{F} an algebra for \coq{F} over the carrier
\coq{A} is a map of the following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}

\dcas{The writing below is confusing, but I am not sure how to fix it ... Can't
we start with the list-generating functor? The monoid example is great, but as 
soon as we move to talking about least-fixed points, it feels as if we'd better
go straight into F X = unit + A * X}
Now, the \emph{easiest} and most effective way to turn any type $A$ into a
monoid is to define it as the least solution to the equation
\begin{coqcode}
Free A = A + unit + Free A * Free A
\end{coqcode}
with \coq{Free : Type -> Type}. In Coq, this is the least fixed-point of the
functor \coq{unit + A + X * X} on \coq{X}.  This is isomorphic to the type of
lists over \coq{A} hence it is the least fixed-point for the
functor \coq{unit + A * X} as well. This is the least ``set'' generated by the
constructors \coq{Empty : unit -> List A} and \coq{Cons : A * List A -> List A}.
These constructors give an algebra \coq{LFix_in : unit + A * List A -> List A}
which has and inverse which we call \coq{LFix_out}.

We now formalise this idea using container extensions.  For a container
extension a data type is formalised as an inductive data type parametrised by
the functor represented by \coq{F}:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
% To generalise recursion schemes we abstract the shape of the data using functors.
% Assuming a base functor \coq{F} a catamorphism is then the unique
% solution to the equation
% \begin{minted}{coq}
% cata alg = alg . fmap (cata alg) . InOp
% \end{minted}
% which does precisely what we described in the previous paragraph.
%
% where $F$ is a \emph{base functor}, $Ff$ is the \emph{functorial action} which
% can be thought of the program $\operator{fmap}\; f$, $a$ is a program of type
% $F A \to A$ (an $F$-algebra),  $c$ is a program of type $C \to FC$ (a
% $F$-coalgebra), $\InOp$ is the isomoprhism on the least fixed-point of $F$,
% namely $\InOp : \mu F \to F\mu F$, and $\OutOp$ is the isomorphism on the
% greatest fixed-point of $F$, namely $F \nu F \to \nu F$. When the unique
% solutions to the above equations exist we denote them as $\cata{a}$ and
% $\ana{c}$ respectively.
%
%%\dcas{Boring stuff}
%The definition of \coq{LFix} in terms of \coq{App} requires us to define our
%own custom induction scheme for initial algebras:
%\begin{coqcode}
%Lemma LFix_ind [P : LFix -> Prop]
%  : (forall x : App F LFix,
%        (forall e : Pos (shape x), P (cont x e)) ->
%        P (LFix_in x))
%    -> forall l : LFix, P l.
%\end{coqcode}
We define \coq{LFix} as a setoid, where its equivalence relation can be
described as the least fixed point of \coq{AppR}.  We define smart constructors
for the isomorphism of least fixed points as respectful morphisms.
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F                l_out : LFix F ~> App F (LFix F)
\end{coqcode}

The most simple and well-known recursion scheme is the \emph{fold} or
\emph{catamorphism} which consumes data. First a fold \emph{destructs} a data
type using \coq{LFix_out}, then calls itself recursively and then folds the
results of the recursive call using an \coq{F}-algebra.
\begin{coqcode}
  Definition cata_f `{F : Cont Sh P} (alg : Alg F A) : LFix -> A
:= fix f (x : LFix) := match x with
  | LFix_in ax => let (sx, kx) := ax in alg (MkApp sx (fun e => f (kx e))) end.
\end{coqcode}
It is easy to show that this function is a respectful morphism of
\coq{F}-algebras. In fact, it is possible to define it as a map of the
following type:
\begin{coqcode}
  cata : forall `{F : Cont Sh P} `{setoid A}, Alg F A ~> LFix ~> A
\end{coqcode}
We prove that catamorphisms satisfy the universal law, that is that there is a
unique morphism of \coq{F}-algebras from the least fixed-point of \coq{F}:
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}

\subsection{Coalgebras and Anamorphisms}
\label{sec:coalg}
While folds \emph{destruct} data, their dual, the unfolds, \emph{construct}
data. In particular, given a seed value and an \emph{observation}, the unfold
corecursively  ``runs'' the observations starting from the initial seed.
% Consider the transition system below composed by three states \coq{x1}, \coq{x2}
% and \coq{x3} with the transitions labelled by the set of labels \coq{\{a, b\}} and define the map \coq{c} such that it implements the transition system below
% \[\begin{tikzcd}
%     {x_1} & {x_2} & {x_3} \arrow["b"', loop, distance=2em, in=35, out=325]
% 	\arrow["a", from=1-1, to=1-2]
% 	\arrow["b", from=1-2, to=1-3]
% \end{tikzcd}\]
The seed function can be modelled by a coalgebra. For example, for a type of
states \coq{X} and an alphabet type \coq{L} we can define a (labelled)
transition system (LTS) on \coq{X} as a function \coq{c : X -> L * X}
implementing the \emph{transition} map. In particular, for a
state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.
Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
carrier \coq{X} is a map \coq{X -> F X}. In the previous example \coq{c} is
an \coq{F}-coalgebra on the carrier \coq{X} with \coq{F X = L * X}.
In our development we use the following notation for coalgebras:
\begin{coqcode}
Notation Coalg F A := (A ~> App F A).
\end{coqcode}
Now an anamorphism takes such a seed function and runs it while collecting all
the observable behaviours. In order to do so we define the type of infinite
streams of labels which is the greatest fixed-point for the functor \coq{A * X}.
We define this in Coq as follows:
\begin{coqcode}
CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
\end{coqcode}
where \coq{GFix_in}  and \coq{GFix_out} are the two inverses witnessing the
isomorphism. Similarly to \coq{LFix}, \coq{GFix} is defined as a setoid, with
an equivalence relation that is the greatest fixpoint of \coq{AppR}.
Additionally, we define smart constructors for the isomorphism of greatest
fixed points
\begin{coqcode}
g_in : App F (GFix F) ~> GFix F                g_out : GFix F ~> App F (GFix F)
\end{coqcode}
\dcas{The explanation below is weird ... We have not defined 'ana', but state
their universal property?}
What witnesses the fact that this is the greatest fixed-point is the universal
property of the anamorphism. For a given \coq{F}-coalgebra this is the unique
map that runs the seed function and composes the information from the first step
\begin{coqcode}
Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
: f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
In our Coq formalisation, we define anamorphisms as the
unique \coq{F}-coalgebra homomorphisms, that are maps of the following type:
\begin{coqcode}
ana : forall `{F : Cont Sh P} `{setoid A}, Coalg F A ~> A ~> GFix F
\end{coqcode}

\subsection{Recursive Coalgebras and Hylomorphism}
\label{sec:rec-coalgebras}
Hylomorphisms capture the concept of \emph{divide-and-conquer} algorithms.
First the input is destructured (\emph{divide}) in smaller parts. This
operation may or may not be structural. Next the results from the subparts are
computed recursively and finally the results are composed together
(\emph{conquer}) by means of an algebra.

@@

\dcas{Marco, shouldn't the below be "given an F-algebra and a recursive
F-coalgebra"? We contradict ourselves below where we say that the hylomorphism
equation does not necessarily have unique solutions, or solutions at all. As
far as I understand, these are also called hylomorphisms. E.g. one can define a
general fixpoint combinator as a hylomorphism, which does not have a unique
solution}
Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
unique solution to the equation
\begin{equation}
  \label{eq:hylo}
  f = a \circ F\; f \circ c
\end{equation}
However this solution does not necessarily exist for an arbitrary pair of
algebras and coalgebras. In fact, this definition cannot be accepted by Coq,
since there is no structurally smaller argument to do recursion on and there is
no corecursive definition that is guaranteed to be syntactically guarded -- as
required by Coq -- due to the use of an arbitrary algebra $a$. Furthermore,
solutions to the hylomorphism equation are \emph{not} guaranteed to be unique,
or to even exist. To address these issues, we mechanise \emph{recursive
hylomorphisms} which are guaranteed to have a unique solution to the
hylomorphism equation. These are hylomoprhisms where the coalgebra is
\emph{recursive}, i.e. coalgebras that terminate on all inputs. The following
predicate captures that a coalgebra terminates on an input: 
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra 
\coq{c : Coalg F A} such that \coq{forall x, RecF c x}.
Recursive hylomorphisms are implemented in Coq recursively on the structure of
the proof for the type (\coq{RecF}) as follows:
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A) 
  : forall (x : A), RecF c x -> B := fix f x H
  := match c x as h0 return (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{coqcode}
We use \coq{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls. As we did with catamorphisms and anamorphisms, we
prove that \coq{hylo_def} is respectful, and use this proof to build the
corresponding higher-order proper morphism:
\begin{coqcode}
  hylo : forall `{F : Cont Sh P} `{setoid A} `{setoid B}, 
    Alg F B ~> RCoalg F A ~> A ~> B
\end{coqcode}
Finally, we show that recursive hylomorphisms are the unique solution to the
hylomorphism equation.
\begin{coqcode}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{coqcode}
Hylomorphisms fusion falls out from the universal property:
\begin{coqcode}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
\end{coqcode}
% Qed.
%   rewrite hylo_univ, fmap_comp, compA, <- E2,
%           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
